---
title: "NCBN Thermal Report"
execute:
  message: false
  warning: false
editor options: 
  markdown:
    wrap: none
output: 
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    fig_caption: true
    number_sections: false
    df_print: kable
    theme: journal
    css: journalnps.min.css
    includes:
        before_body:
          - header.html
date: "2026-05-05"
params:
  park: "CACO" #change to FIIS or CACO when knitting for park specific report
  park_name: "Cape Cod National Seashore"
---

```{r setup, include=FALSE}

# Libraries (load once)
library(tidyverse)
library(readxl)
library(ggrepel)
library(glue)
library(fs)
library(scales)   # for label_number
library(here) # use to load data from relative paths

# -------------------------------
# 0) GLOBAL STYLE + PALETTE
# -------------------------------
theme_set(theme_classic(base_size = 14))

# Locked legend order for Park_Code
park_levels <- c("ASIS", "FIIS", "CACO")

# Consistent colors and shapes everywhere
park_pal    <- c("ASIS"="#1f77b4","FIIS"="#ff7f00","CACO"="#2ca02c")
park_shapes <- c("ASIS"=16,"FIIS"=17,"CACO"=15)

# Accent colors for thresholds/extreme markers (not mapped to Park_Code)
grey_grid     <- "grey40"
accent_red    <- "#d62728"
accent_blue   <- "#1f78b4"
accent_purple <- "#8b3fb9"

# Reusable scales—use these in every plot that maps color/fill/shape to Park_Code
scale_park_color <- function(...) scale_color_manual(values = park_pal, breaks = park_levels, ...)
scale_park_fill  <- function(...) scale_fill_manual(values  = park_pal, breaks = park_levels, ...)
scale_park_shape <- function(...) scale_shape_manual(values = park_shapes, breaks = park_levels, ...)

```

```{r Load Data, include= FALSE}
dat <- read_excel(here::here("data", "Site_Summary_Master_b.xlsx"), sheet = "Site_Year") %>%
  rename_with(~ gsub(" ", "_", .)) %>%
  mutate(
    Park_Code              = toupper(Park_Code),
    Park_Code              = factor(Park_Code, levels = park_levels), # lock legend order
    Mean_HDH_daily         = suppressWarnings(as.numeric(Mean_HDH_daily)),
    Mean_CDH_daily         = suppressWarnings(as.numeric(Mean_CDH_daily)),
    `%_abv_thres`          = suppressWarnings(as.numeric(`%_abv_thres`)),
    Number_warming_events  = suppressWarnings(as.numeric(Number_warming_events)),
    Year                   = suppressWarnings(as.integer(Year)),
    Transect               = toupper(gsub("[^A-Za-z]", "", trimws(Transect)))
  )


```

```{r helper functions, include=FALSE}
# Site-year (avg across transects) -> Site-level (avg across years)
make_site_level <- function(df, metrics = c("Mean_HDH_daily","Mean_CDH_daily","%_abv_thres")) {
  df %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(
      HDH_year     = if ("Mean_HDH_daily" %in% metrics) mean(Mean_HDH_daily, na.rm = TRUE) else NA_real_,
      CDH_year     = if ("Mean_CDH_daily" %in% metrics) mean(Mean_CDH_daily, na.rm = TRUE) else NA_real_,
      pct_hot_year = if ("%_abv_thres"    %in% metrics) mean(`%_abv_thres`,    na.rm = TRUE) else NA_real_,
      .groups = "drop"
    ) %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      mean_HDH = mean(HDH_year,     na.rm = TRUE),
      mean_CDH = mean(CDH_year,     na.rm = TRUE),
      pct_hot  = mean(pct_hot_year, na.rm = TRUE),
      n_years  = dplyr::n_distinct(Year),
      .groups  = "drop"
    ) %>%
    dplyr::mutate(Label = paste0(Site_Name, " (n=", n_years, ")"))
}

# Flag extreme years per site (e.g., 90th percentile of metric within each site)
flag_extreme_by_site <- function(df, value_col, prob = 0.9, flag_name = "Extreme") {
  df %>%
    dplyr::group_by(Site_Name) %>%
    dplyr::mutate(
      thresh = stats::quantile(.data[[value_col]], prob, na.rm = TRUE),
      "{flag_name}" := .data[[value_col]] >= .data[["thresh"]]
    ) %>%
    dplyr::ungroup()
  
}

# Number of monitoring years per park
years_per_park <- function(df) {
  df %>%
    dplyr::group_by(Park_Code) %>%
    dplyr::summarise(
      n_years    = dplyr::n_distinct(Year),
      first_year = min(Year, na.rm = TRUE),
      last_year  = max(Year, na.rm = TRUE),
      .groups    = "drop"
    )
}

# Per-park site summary
site_summary <- function(df) {
  df %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      n_years = dplyr::n_distinct(Year),
      .groups = "drop"
    )
}

```

```{r filter to park, include= FALSE}
park_code <- params$park
dat_pk <- dat %>% filter(park_code == park_code)
```

```{r Per Park Plotting, include= FALSE}
# A) High Heat Years (HDH) by site (facetted), one park
plot_hdh_extreme_by_site <- function(df, park_code) {
  site_year <- df %>%
    dplyr::filter(Park_Code == park_code) %>%
    dplyr::group_by(Site_Name, Year) %>%
    dplyr::summarise(Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE), .groups = "drop") %>%
    flag_extreme_by_site(value_col = "Mean_HDH", prob = 0.9, flag_name = "Extreme")
  
  ggplot(site_year, aes(x = Year, y = Mean_HDH)) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(size = 2) +
    geom_point(data = subset(site_year, Extreme), color = accent_red, size = 3.5) +
    geom_text_repel(
      data = subset(site_year, Extreme),
      aes(label = Year), color = accent_red, size = 4, show.legend = FALSE,
      nudge_x = 0.4, nudge_y = 0.6, hjust = 0, direction = "y",
      box.padding = 0.3, point.padding = 0.2
    ) +
    geom_hline(aes(yintercept = thresh), linetype = "dashed", color = accent_red) +
    facet_wrap(~ Site_Name, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = label_number(accuracy = 1),
                       expand = expansion(mult = c(0.1, 0.2))) +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE), max(site_year$Year, na.rm = TRUE), by = 2)) +
    labs(
      title    = glue("High Heat Years by Site ({park_code})"),
      subtitle = "Red dashed line = site-specific threshold; red points = extreme heat years",
      x = "Year", y = "Mean Heating Degree Hours per Day"
    )
}

# B) Reduced Cooling Years (CDH) by site (facetted), one park
plot_cdh_lowcool_by_site <- function(df, park_code) {
  site_year <- df %>%
    dplyr::filter(Park_Code == park_code) %>%
    dplyr::group_by(Site_Name, Year) %>%
    dplyr::summarise(Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE), .groups = "drop") %>%
    flag_extreme_by_site(value_col = "Mean_CDH", prob = 0.9, flag_name = "Low_Cooling")
  
  ggplot(site_year, aes(x = Year, y = Mean_CDH)) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(size = 2) +
    geom_point(data = subset(site_year, Low_Cooling), color = accent_blue, size = 3.5) +
    geom_text_repel(
      data = subset(site_year, Low_Cooling),
      aes(label = Year), color = accent_blue, size = 4, show.legend = FALSE,
      nudge_x = 0.4, nudge_y = 0.6, hjust = 0, direction = "y",
      box.padding = 0.3, point.padding = 0.2
    ) +
    geom_hline(aes(yintercept = thresh), linetype = "dashed", color = accent_blue) +
    facet_wrap(~ Site_Name, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = label_number(accuracy = 1),
                       expand = expansion(mult = c(0.1, 0.2))) +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE), max(site_year$Year, na.rm = TRUE), by = 2)) +
    labs(
      title    = glue("Reduced Cooling Years by Site ({park_code})"),
      subtitle = "Blue dashed line = site-specific threshold; blue points = low cooling years",
      x = "Year", y = "Mean Cooling Degree Hours per Day"
    )
}

# C) Double-Stress years (HDH high & CDH high), one park
plot_double_stress_by_site <- function(df, park_code) {
  site_year <- df %>%
    dplyr::filter(Park_Code == park_code) %>%
    dplyr::group_by(Site_Name, Year) %>%
    dplyr::summarise(
      Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE),
      Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Site_Name) %>%
    dplyr::mutate(
      hdh_thresh = stats::quantile(Mean_HDH, 0.9, na.rm = TRUE),
      cdh_thresh = stats::quantile(Mean_CDH, 0.9, na.rm = TRUE),
      High_Heat   = Mean_HDH >= hdh_thresh,
      Low_Cooling = Mean_CDH >= cdh_thresh,
      Double_Stress = High_Heat & Low_Cooling
    ) %>% dplyr::ungroup()
  
  ggplot(site_year, aes(x = Year, y = Mean_HDH)) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(size = 2) +
    geom_point(data = subset(site_year, Double_Stress), color = accent_purple, size = 4) +
    geom_text_repel(
      data = subset(site_year, Double_Stress),
      aes(label = Year), color = accent_purple, size = 4, show.legend = FALSE,
      nudge_x = 0.4, nudge_y = 0.6, hjust = 0, direction = "y",
      box.padding = 0.3, point.padding = 0.2
    ) +
    geom_hline(aes(yintercept = hdh_thresh), linetype = "dashed", color = accent_red) +
    facet_wrap(~ Site_Name, ncol = 1, scales = "free_y") +
    scale_y_continuous(labels = label_number(accuracy = 1),
                       expand = expansion(mult = c(0.1, 0.2))) +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE), max(site_year$Year, na.rm = TRUE), by = 2)) +
    labs(
      title    = glue("Thermal Stress Years by Site ({park_code})"),
      subtitle = "Purple points = years with high heat AND reduced cooling",
      x = "Year", y = "Mean Heating Degree Hours per Day"
    )
}
```

```{r Regional PLotting Functions, include= FALSE}
# D) Site-level thermal stress across years (lines colored by Park_Code)
plot_sitelevel_thermal_stress_lines <- function(df) {
  site_year <- df %>%
    dplyr::filter(Park_Code %in% park_levels) %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE), .groups = "drop")
  
  threshold <- stats::quantile(site_year$Mean_HDH, 0.9, na.rm = TRUE)
  
  site_labels <- site_year %>%
    dplyr::group_by(Site_Name) %>%
    dplyr::filter(Year == max(Year, na.rm = TRUE)) %>%
    dplyr::ungroup()
  
  ggplot(site_year,
         aes(x = Year, y = Mean_HDH, group = Site_Name, color = Park_Code)) +
    geom_line(alpha = 0.7, linewidth = 1) +
    geom_point(size = 2) +
    geom_text_repel(
      data = site_labels,
      aes(label = Site_Name),
      size = 3.5, direction = "y", hjust = 0,
      nudge_x = 1, box.padding = 0.5, point.padding = 0.3,
      segment.color = "grey60", show.legend = FALSE
    ) +
    geom_point(
      data = subset(site_year, Mean_HDH >= threshold),
      color = accent_red, size = 3
    ) +
    geom_text_repel(
      data = subset(site_year, Mean_HDH >= threshold),
      aes(label = Year),
      size = 3, color = accent_red, show.legend = FALSE
    ) +
    geom_hline(yintercept = threshold, linetype = "dashed", color = accent_red) +
    scale_park_color() +
    scale_x_continuous(breaks = seq(min(site_year$Year, na.rm = TRUE),
                                    max(site_year$Year, na.rm = TRUE), by = 2)) +
    expand_limits(x = max(site_year$Year, na.rm = TRUE) + 8) +
    labs(
      title    = "Site-Level Thermal Stress Across Years",
      subtitle = "Lines = sites; labels show site identity and extreme heat years",
      x = "Year", y = "Mean Heating Degree Hours per Day",
      color = "Park"
    )
}

# E) Thermal stress vs cooling (site-level scatter with size = % hot)
plot_stress_vs_cooling <- function(df) {
  site_level <- make_site_level(df, metrics = c("Mean_HDH_daily","Mean_CDH_daily","%_abv_thres"))
  median_HDH <- 8.8
  median_CDH <- -94.9
  
  x_low  <- median_HDH * 0.5
  x_high <- max(site_level$mean_HDH, na.rm = TRUE) * 0.8
  y_low  <- min(site_level$mean_CDH, na.rm = TRUE) * 0.8
  y_high <- median_CDH * 0.8
  
  ggplot(site_level, aes(x = mean_HDH, y = mean_CDH, color = Park_Code, size = pct_hot)) +
    geom_point(alpha = 0.9) +
    geom_text_repel(aes(label = Label), size = 4) +
    geom_vline(xintercept = median_HDH, linetype = "dashed", color = grey_grid) +
    geom_hline(yintercept = median_CDH, linetype = "dashed", color = grey_grid) +
    annotate("text", x = x_low,  y = y_high, label = "Stable Systems",          size = 4.5, fontface = "bold") +
    annotate("text", x = x_high, y = y_high, label = "Chronic Heat Stress",     size = 4.5, fontface = "bold") +
    annotate("text", x = x_low,  y = y_low,  label = "Thermal Refugia",         size = 4.5, fontface = "bold") +
    annotate("text", x = x_high, y = y_low,  label = "Stress Buffered Systems", size = 4.5, fontface = "bold") +
    expand_limits(x = max(site_level$mean_HDH, na.rm = TRUE) * 1.2,
                  y = min(site_level$mean_CDH, na.rm = TRUE) * 1.2) +
    scale_park_color() +
    guides(color = guide_legend(override.aes = list(alpha = 1, size = 4))) +
    labs(
      title = "Thermal Stress vs Cooling Relief Across Coastal Monitoring Sites",
      subtitle = "Site-level averages (transects averaged within year, then across years)",
      x = "Mean Heating Degree Hours per Day (Thermal Stress)",
      y = "Mean Cooling Degree Hours per Day (Thermal Recovery)",
      color = "Park",
      size  = "% Time Above Threshold"
    )
}

# F) Thermal balance (exposure vs cooling, site-level)
plot_thermal_balance <- function(df) {
  thermal_balance <- df %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(
      Exposure_year = mean(`%_abv_thres`, na.rm = TRUE),
      Cooling_year  = mean(Mean_CDH_daily,  na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      MeanExposure = mean(Exposure_year, na.rm = TRUE),
      MeanCooling  = mean(Cooling_year,  na.rm = TRUE),
      n_years      = dplyr::n_distinct(Year),
      .groups      = "drop"
    ) %>%
    dplyr::mutate(Label = paste0(Site_Name, " (n=", n_years, ")"))
  
  ggplot(thermal_balance, aes(x = MeanExposure, y = MeanCooling, color = Park_Code)) +
    geom_point(size = 4, alpha = 0.9) +
    geom_text_repel(aes(label = Label), size = 3) +
    scale_park_color() +
    labs(
      title    = "Thermal Balance Across Coastal Monitoring Sites",
      subtitle = "Exposure vs cooling using consistent site-level averaging",
      x = "Mean % Time Above Threshold (Thermal Exposure)",
      y = "Mean Cooling Degree Hours (Thermal Recovery)",
      color = "Park"
    )
}

# G) Heat Event Structure (site-level; optional interpretive quadrants)
plot_heat_event_structure_site <- function(df) {
  site_level_events <- df %>%
    dplyr::group_by(Park_Code, Site_Name, Year) %>%
    dplyr::summarise(
      Events_year   = mean(Number_warming_events, na.rm = TRUE),
      Exposure_year = mean(`%_abv_thres`,        na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::group_by(Park_Code, Site_Name) %>%
    dplyr::summarise(
      MeanEvents   = mean(Events_year,   na.rm = TRUE),
      MeanExposure = mean(Exposure_year, na.rm = TRUE),
      n_years      = dplyr::n_distinct(Year),
      .groups      = "drop"
    ) %>%
    dplyr::mutate(Label = paste0(Site_Name, " (n=", n_years, ")"))
  
  cluster_df <- site_level_events %>% # create df with only Events and Exposure for fitting k-means
    mutate(id = paste0(Park_Code, "_", Site_Name)) %>%
    column_to_rownames(var = "id") %>%
    ungroup() %>%
    dplyr::select(MeanEvents, MeanExposure)
  
  km_fit <- kmeans(cluster_df, centers = 4, nstart = 2) # fit k-means clusters
  event_split <- mean(km_fit$centers[1:4])
  exposure_split <- mean(km_fit$centers[5:8])
  
  ggplot(site_level_events, aes(x = MeanEvents, y = MeanExposure, color = Park_Code, shape = Park_Code)) +
    geom_point(size = 4.5) +
    geom_text_repel(aes(label = Label), size = 3, max.overlaps = 30) +
    geom_vline(xintercept = event_split,   linetype = "dashed", color = grey_grid) +
    geom_hline(yintercept = exposure_split,linetype = "dashed", color = grey_grid) +
    annotate("text", x = event_split * 0.45, y = max(site_level_events$MeanExposure, na.rm = TRUE) * 0.9,
             label = "Episodic Heat Waves", size = 4) +
    annotate("text", x = max(site_level_events$MeanEvents,   na.rm = TRUE) * 0.75, y = max(site_level_events$MeanExposure, na.rm = TRUE) * 0.9,
             label = "Highly Stressed", size = 4) +
    annotate("text", x = max(site_level_events$MeanEvents,   na.rm = TRUE) * 0.75, y = exposure_split * 0.5,
             label = "Chronic Warming", size = 4) +
    annotate("text", x = event_split * 0.45, y = exposure_split * 0.45,
             label = "Low Exposure", size = 4) +
    scale_park_color() +
    scale_park_shape() +
    labs(
      title    = "Heat Event Structure Across Coastal Parks",
      subtitle = "Points represent site-level averages across monitoring years",
      x = "Mean Number of Warming Events (Frequency)",
      y = "Mean % Time Above Threshold (Duration)",
      color = "Park", shape = "Park"
    )
}

# (Optional) H) Thermal balance at TRANSECT level (no interpretation grids)
plot_thermal_balance_transect <- function(df) {
  thermal_balance <- df %>%
    dplyr::filter(Park_Code %in% park_levels) %>%
    dplyr::group_by(Park_Code, Site_Name, Transect) %>%
    dplyr::summarise(
      MeanExposure = mean(`%_abv_thres`,    na.rm = TRUE),
      MeanCooling  = mean(Mean_CDH_daily,   na.rm = TRUE),
      n_years      = dplyr::n(),
      .groups      = "drop"
    ) %>%
    dplyr::mutate(Label = paste(Site_Name, Transect))
  
  ggplot(thermal_balance, aes(x = MeanExposure, y = MeanCooling, color = Park_Code)) +
    geom_point(size = 4) +
    geom_text_repel(aes(label = Label), size = 3) +
    scale_park_color() +
    labs(
      title = "Thermal Balance Across Parks (Transect Level)",
      x = "Mean % Time Above Threshold",
      y = "Mean Cooling Degree Hours",
      color = "Park"
    )
}

```


## Eelgrass Importance and Vulnerability

Eelgrass (Zostera marina) is a foundational marine plant that supports healthy coastal ecosystems. It improves water quality, stabilizes sediments, provides habitat for fish and invertebrates, and acts as a globally important carbon sink (Novak et al. 2023). When eelgrass meadows decline, the carbon they store can be released back into the water column (Berger et al. 2024). Because eelgrass responds quickly to changes in its environment, it is often considered a “canary in the coal mine” for coastal health (Orth et al. 2006). Its ecological significance is one reason the Northeast Coastal and Barrier Network (NCBN) monitors seagrass conditions over time, especially as concerns rise about the effects of warming waters.

In recent decades, eelgrass along the Northeast U.S. coast has experienced water temperatures that approach or exceed known physiological limits. When water becomes too warm, eelgrass respiration increases faster than photosynthesis, creating stress on the plants. Studies show that temperatures above 25 °C increase the likelihood of eelgrass mortality, and sustained exposure above 30 °C has led to complete die offs in areas such as Chesapeake Bay (Jarvis & Moore 2010; Plaisted et al. 2022, Novak et al. 2023). Long term temperature records from New Hampshire to Maryland indicate that warmer summers are consistently followed by lower eelgrass presence in the next year (Plaisted et al. 2022). Temperature stress is also more severe when combined with poor water clarity, low light, or high nutrient levels. Repeated hot summers without adequate recovery time can lead to large scale declines.

These patterns align with broader climate driven trends. Marine heatwaves, periods when water temperatures exceed the 90th percentile of a 30 year baseline, have become more frequent and more intense (Berger et al. 2024). Such events have already affected corals, kelp forests, and seagrasses worldwide. Research shows that eelgrass meadows experiencing both high heat and little cooling tend to lose shoot density and canopy height, while sites that receive periodic cooling are better able to recover between heat events (Berger et al. 2024).

Other marine systems use cumulative heat stress metrics to understand these risks. NOAA’s Degree Heating Weeks (DHW) index, for example, tracks how much and how long temperatures exceed thresholds that lead to coral bleaching and mortality (Skirving et al. 2020). These tools emphasize that it is the accumulation of heat over time, not single temperature spikes, that often drives ecosystem decline. This same concept is directly relevant to eelgrass.

Calculating cumulative stress frameworks requires long term temperature datasets and NCBN continuous monitoring program does exactly that. To better understand how warming affects eelgrass within the Northeast Coastal and Barrier Network, we used long term temperature datasets to:

1.	Identify years with extreme heat and low cooling,
2.	Determine where these conditions overlapped to create compounded “stress years,” and
3.	Compare patterns across parks to understand regional vulnerability.


# Methods

## Protocol

The Northeast Coastal and Barrier Network uses a standardized Estuarine Nutrient Enrichment monitoring protocol (Kopp and Neckles, 2009) that includes seagrass condition monitoring (SOP 9). Seagrass condition is characterized using measures of population and shoot structure to track how meadows change. This protocol was developed from the international SeagrassNet program that provides framework for standardzied seagrass monitoring worldwide. 

Within the network, three coastal parks, Assateague Island (ASIS), Fire Island (FIIS), and Cape Cod (CACO) implement the seagrass monitoring protocol.Seagrass condition is monitored across three permanent 50-m transects that run parallel to shore at a shallow, mid-depth, and deep depth. Temperature loggers are attached to each transect to record continuous water temperature data  collected once a year. These longterm records give us on the ground data to observe the thermal environment expereinced by eelgrass meadows. 

For this analysis, we used the continuous water temperature data and focused on the months from May-September to hone in on when thermal stress likely occurs. To ensure data quality in each sampling year, we only include years with 80 percentage data coverage and with at least 10 days of data in three of five months.From this dataset we examined patterns of thermal stress and cooling dynamics with the goal to compare the conditions at  `r params$park` with those at  other long-term National Park Service monitoring sites. 


```{r,echo=FALSE}
library(dplyr)
library(knitr)

# Use your existing site_summary() and filter to the selected park
site_sum <- site_summary(dat) %>%
  filter(Park_Code == params$park) %>%
  arrange(Site_Name)

# Clean, simple table using only knitr::kable()
kable(
  site_sum,
  col.names = c("Park Code", "Site Name", "Years of Data"),
  caption = paste0(
    "Available datasets for each site in ",
    params$park,
    " after data filters were applied."
  )
)
```


```{r site-info, include=FALSE}
# Sites NOT considered long-term monitoring
excluded_sites <- c("EAST HARBOR", "PROVINCETOWN HARBOR")

# All long-term monitoring sites for this park
longterm_sites <- site_summary(dat) %>%
  filter(Park_Code == params$park,
         !toupper(Site_Name) %in% excluded_sites) %>%
  pull(Site_Name)

# Formatting for narrative (comma-separated list)
longterm_sites_text <- paste(longterm_sites, collapse = ", ")

# Years of monitoring for the park (unchanged)
n_years <- years_per_park(dat) %>%
  filter(Park_Code == params$park) %>%
  pull(n_years)

first_year <- years_per_park(dat) %>%
  filter(Park_Code == params$park) %>%
  pull(first_year)

last_year <- years_per_park(dat) %>%
  filter(Park_Code == params$park) %>%
  pull(last_year)

```

## Novel Thermal Metrics for Eelgrass

To describe long-term thermal conditions in eelgrass meadows, we calculate two metrics from our continuous water temperature record: **heating degree hours (HDH)** and **cooling degree hours (CDH)**. Following the methods outlined by Berger et al. 2024, these metrics capture not just how hot the water gets, but how long water temperatures remain above or below a biological threshold.

For eelgrass, prolonged exposure to high temperatures can lead to physiological stress where respiration outpaces photosynthesis. We applied a threshold of 25 °C for northern sites and 28.6 °C for southern sites, values associated with known limits in eelgrass thermal tolerance.

Together, HDH and CDH show how often heat stress events occur and how well the area recovers from them. Looking at both gives us a clearer picture of the factors that shape eelgrass resilience. 

```{r concepts, echo=FALSE, warning=FALSE, fig.show='hold', fig.width=4, fig.width=4}
### HDH and CDH Concept##

library(tidyverse)

df <- tibble(
  time = 1:100,
  temp = 25 + sin(time/10)*3
)

threshold <- 25

ggplot(df, aes(x = time, y = temp)) +
  geom_line(size = 1.2) +
  
  geom_hline(yintercept = threshold, linetype = "dashed") +
  
  geom_ribbon(
    data = df %>% filter(temp > threshold),
    aes(ymin = threshold, ymax = temp),
    fill = "red",
    alpha = 0.4
  ) +
  
  labs(title = "HDH = Heat Above Threshold") +
  theme_classic()
ggplot(df, aes(x = time, y = temp)) +
  geom_line(size = 1.2) +
  
  geom_hline(yintercept = threshold, linetype = "dashed") +
  
  geom_ribbon(
    data = df %>% filter(temp < threshold),
    aes(ymin = temp, ymax = threshold),
    fill = "blue",
    alpha = 0.4
  ) +
  
  labs(title = "CDH = Cooling Below Threshold") +
  theme_classic()


```

# Results

At `r params$park`, the long‑term monitoring site(s) include **`r longterm_sites_text`**, with `r n_years` years of data collected between `r first_year` and `r last_year`. Additional points of interest were included in the analysis to provide broader ecological context but are not included in long‑term monitoring protocol.

## Extreme Heat Years
At `r params$park`, we identified extreme heat years for each long-term monitoring site by calculating the 90th percentiles of heating degree hours (HDH) as a site-specific threshold. The red points indicated years where the mean HDH exceeded this threshold and are considered "extreme heat years" for that site. These years represent years when the eelgrass meadow at these sites experiences the more intense and prolonged thermal stress. The table below lists years at each moitoring locations when heat degree around were exceptionally high (i.e, HDH above the 90th percentile for that site).


```{r plot-hdh, fig.width=10, fig.height=8, fig.cap="<small>Extreme heat years (red points) are years when mean HDH exceeded the site-specific 90th percentile threshold. Each panel shows a long-term monitoring site in `r params$park`.</small>", echo=FALSE}
plot_hdh_extreme_by_site(dat, park_code)
```

```{r hdh-extreme-summary, echo=FALSE, results='asis'}
library(dplyr)
library(knitr)

# Prepare site-year data for the selected park
site_year <- dat %>%
  filter(Park_Code == params$park) %>%
  group_by(Site_Name, Year) %>%
  summarise(Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE), .groups = "drop")

# Use your helper to flag extreme years (90th percentile)
site_year_extreme <- flag_extreme_by_site(site_year, value_col = "Mean_HDH", prob = 0.9, flag_name = "Extreme") %>%
  filter(Extreme) %>%
  arrange(Site_Name, Year)

# Output summary table
if (nrow(site_year_extreme) > 0) {
  site_year_extreme %>%
    group_by(Site_Name) %>%
    summarise(
      `Extreme heat years` = paste(Year, collapse = ", ")
    ) %>%
    kable(
      caption = paste0("Extreme heat years (HDH above 90th percentile) at each long-term site in ", params$park, "."),
      col.names = c("Site", "Extreme heat years")
    )
} else {
  cat("No extreme heat years were detected at the long-term monitoring sites in this park.")
}

```


## Reduced Cooling Years

On the flip side of heat degree hours, we can look at the mean cooling degree hours across `r params$park`. In this context, more negative CDH values represents greater cooling below the temperature threshold at each site. This represents periods of potential relief for eelgrass meadows.

The table below lists the years at each long term monitoring site when cooling degree hours were exceptionally low (i.e, CDH above the 90th percentile for that site). Extreme low cooling years give us insight when thermal relief was lowest that could limit recovery from thermal stress events.



```{r plot-cdh, fig.width=10, fig.height=8, fig.cap="<small>Low cooling years (blue points) are years when mean CDH exceeded the site-specific 90th percentile threshold. Each panel shows a long-term monitoring site in `r params$park`.</small>", echo=FALSE}
plot_cdh_lowcool_by_site(dat, park_code)
```


```{r cdh-extreme-summary, echo=FALSE, results='asis'}
library(dplyr)
library(knitr)

# Prepare site-year data for the selected park
site_year_cdh <- dat %>%
  filter(Park_Code == params$park) %>%
  group_by(Site_Name, Year) %>%
  summarise(Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE), .groups = "drop")

# Use your helper to flag extreme low cooling years (90th percentile)
site_year_cdh_extreme <- flag_extreme_by_site(site_year_cdh, value_col = "Mean_CDH", prob = 0.9, flag_name = "Extreme") %>%
  filter(Extreme) %>%
  arrange(Site_Name, Year)

# Output summary table
if (nrow(site_year_cdh_extreme) > 0) {
  site_year_cdh_extreme %>%
    group_by(Site_Name) %>%
    summarise(
      `Extreme low cooling years` = paste(Year, collapse = ", ")
    ) %>%
    kable(
      caption = paste0("Extreme low cooling years (CDH above 90th percentile) at each long-term site in ", params$park, "."),
      col.names = c("Site", "Extreme low cooling years")
    )
} else {
  cat("No extreme low cooling years were detected at the long-term monitoring sites in this park.")
}
```

## Stress Years (Combined Heat and low Cooling)

DRAFT

Now we look at where the years of extreme heat and low cooling intersect. Compounded we defined these as "stress years" at the site. 
```{r plot-stress, fig.width=10, fig.height=8, fig.cap="<small> Stress years (purple points) are years when mean CDH **and** mean HDH exceeded the site-specific 90th percentile threshold, inidicating compounded stress.</small>", echo=FALSE}
plot_double_stress_by_site(dat,park_code)

```

```{r double-stress-summary, echo=FALSE, results='asis'}
library(dplyr)
library(knitr)

# Summarize double‑stress years (HDH high & CDH high) for each site in the selected park

double_stress <- dat %>%
  filter(Park_Code == params$park) %>%
  group_by(Site_Name, Year) %>%
  summarise(
    Mean_HDH = mean(Mean_HDH_daily, na.rm = TRUE),
    Mean_CDH = mean(Mean_CDH_daily, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Site_Name) %>%
  mutate(
    hdh_thresh = quantile(Mean_HDH, 0.9, na.rm = TRUE),
    cdh_thresh = quantile(Mean_CDH, 0.9, na.rm = TRUE),
    High_Heat   = Mean_HDH >= hdh_thresh,
    Low_Cooling = Mean_CDH >= cdh_thresh,   # high CDH = low cooling
    Double_Stress = High_Heat & Low_Cooling
  ) %>%
  filter(Double_Stress) %>%
  arrange(Site_Name, Year)

if (nrow(double_stress) > 0) {
  double_stress %>%
    group_by(Site_Name) %>%
    summarise(
      Double_stress_years = paste(Year, collapse = ", ")
    ) %>%
    kable(
      caption = paste0(
        "Double‑stress years (high heat & reduced cooling) at each long‑term site in ",
        params$park, "."
      ),
      col.names = c("Site", "Double‑stress years")
    )
} else {
  cat("No double‑stress years were detected at the long‑term monitoring sites in this park.")
}

```

## How does Park Compare Across the Region?

Not all thermal stress at each seagrass monitoring site looks the same. Here we divide the eelgrass meadows into four thermal regimes based on k cluster means:

* **Episodic Heatwave**: Sites with infrequent, but intense heat events
* **Highly Stressed**: Sites with frequent and prolonged heat events
* **Low Exposure**: Sites with infrequent and limited time above thermal threshold
* **Chronic Warming**: Sites with frequent heat events that are not prolonged (pulsing)


```{r regional results, echo=FALSE, fig.width=10, fig.height=6, fig.cap= "<small> Heat event structure across three coastal National Park Service units (CACO, ASIS, FIIS). Each point represents a site‑level average of the mean number of warming events per year (frequency) and the mean percentage of time above the thermal threshold (duration). Dashed lines divide the figure into four interpretive thermal regimes. Labels show the average number of monitoring years contributing to each site‑level estimate.</small>"}
plot_heat_event_structure_site(dat)

```

Across the three coastal NCBN parks, clear regional patterns emerge. At CACO, most sites fall in the category of having fewer and shorter heat‑stress events. An exception is the partially restored East Harbor system, which shows a different pattern. Overall, CACO sites tend to experience heat stress as brief or infrequent pulses rather than long, sustained events.

In contrast, Tingles Island at ASIS experience heat stress much more often, and these events add up throughout the season. This pattern is expected for parks located in warmer parts of the region, where longer and more frequent heat events are common.

FIIS sites fall between these two extremes. Heat‑stress conditions vary widely across the park, depending on where each site is located. For example, Moriches Bay shows short, frequent warming events, while the Great South Bay transect experiences longer or more sustained heat episodes. This mix reflects the diversity of habitats and local conditions across Fire Island.

These results show us that sites across the regional scale experience different heat stress conditions and even sites within the same park respond in different ways. This raises an important questions **how are some of the most chronically stressed sites, like Moriches Bay, Pleasant Bay, and Tingles Island,still able to support eelgrass populations**. 

## What Are the Impact of Cooling?

Now that we've seen sites experience heat stress differently, we can look at relief from heat events. Here we divide the eelgrass into four thermal regimes based on the balance between thermal stress and thermal relief:

* **Stable Systems**: Low heat stress and moderate cooling
* **Chronic Heat Stress**: High heat stress and limited cooling
* **Thermal Refugia**: Low heat stress and strong cooling
* **Stress Buffered Systems**:High heat stress but strong cooling


```{r, echo= FALSE, fig.width=12, fig.height=6, fig.cap="<small> Thermal stress and cooling recovery across three coastal National Park Service units (CACO, ASIS, FIIS). Each point represents a site‑level average of the mean heating degree hours per day (thermal stress) and the mean cooling degree hours per day (thermal recovery). Dashed lines divide the figure into four interpretive recovery regimes. Labels show the average number of monitoring years contributing to each site‑level estimate.</small>",}
plot_stress_vs_cooling(dat)
```

The balance between heat stress events and cooling recovery also varies across the region and within parks. Pleasant Bay and Moriches Bay receive some of the strongest cooling periods among the chronically warm sites, second only to the fully open tidal systems on Cape Cod. This pattern reflects their positions to inlets where tidal flushing brings in cooler ocean water taht helps moderate warmer temperatures and sustain populations of eelgrass. In contrast, sites with more limited flushing, like Great South Bay and East Harbor, show weaker cooling relief consistent with restricted or partially restored flow. These sites match site conditions that no longer support eelgrass. Tingles Island falls between these two outcomes as it receives little cooling, sits between a stable and thermal refugia zone, and supports sparse eelgrass populations. 

# Key Takeaways

DRAFT

# References

Berger AC, Berg P, McGlathery KJ, Aoki LR, Kerns K (2024) Eelgrass meadow response to heat stress. II. Impacts of ocean warming and marine heatwaves measured by novel metrics. Mar Ecol Prog Ser 736:47-62 https://doi.org/10.3354/meps14588

Jarvis, J.C. & Moore, K.A. (2010). The role of the photosynthetic pathway in the acclimation of Zostera marina to light and temperature stress. Journal of Experimental Marine Biology and Ecology, 395, 105–110.

Kopp, B. S., and H. A. Neckles. 2009. A protocol for monitoring estuarine nutrient enrichment in
coastal parks of the National Park Service Northeast Region. Natural Resource Report
NPS/NCBN/NRR—2009/110. National Park Service, Fort Collins, Colorado.

Novak, A. B., Plaisted, H. K., Hughes, Z. J., Mittermayr, A., & Molden, E. (2023). Eelgrass (Zostera marina L.) populations are threatened by high sea-surface temperatures and impaired waters on Nantucket Island, USA. Marine Pollution Bulletin, 197, 115689.

Orth, R. J., Carruthers, T. J., Dennison, W. C., Duarte, C. M., Fourqurean, J. W., Heck, K. L., ... & Williams, S. L. (2006). A global crisis for seagrass ecosystems. Bioscience, 56(12), 987-996.

Plaisted, Holly K., et al. "Influence of rising water temperature on the temperate seagrass species eelgrass (Zostera marina L.) in the Northeast USA." Frontiers in Marine Science 9 (2022): 920699.

Skirving, William, et al. "CoralTemp and the coral reef watch coral bleaching heat stress product suite version 3.1." Remote Sensing 12.23 (2020): 3856.
